Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands

### 2024-12-02 index.qmd
Update index.qmd

![Map of Arizona indicating US Geological Survey (USGS) streamgages used in this study. 8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea_20241126.png){#fig-study-area}

@@

Base flow was calculated using a single-parameter, recursive digital filter technique from [@nathan1990]. This base-flow separation technique is based on a recursive digital filter used in signal analysis that separates high-frequency signals (quickflow) from low-frequency signals (base flow) [@lyne1979]. This technique has been used in [@ARNOLD200021; @santhi2008; @bloomfield2009; @singh2018]. It takes the form of

@@

where $b$ is base flow, $\alpha$ is the filter parameter, $Q$ is the total streamflow, and $k$ is the time step. A filter parameter $\alpha$ of 0.925 was used as in @nathan1990 and @fuka2014ecohydrology. The filter was run three times (forward, backward, forward) to attenuate the base-flow signal. @eckhardt2023 noted that recursive digital filters lack a physical basis, but as the method is easy to automate, objective, and repeatable, it is appropriate for a regional-scale study.

@@

![Mean annual BFI for the period of study (1991-2020) from instrumented stream flow data.](images/BFI_Instrumented_20241126.png){#fig-instrumented-bfi}

@@

| Classification Group         |   n | n_pos | n_neg | perc_pos | perc_neg |
|:-----------------------------|----:|------:|------:|---------:|---------:|
| Climate - Monsoon Dominated  |  94 |     9 |    23 |    0.096 |    0.245 |
| Climate - Snowmelt Dominated | 150 |    19 |    17 |    0.127 |    0.113 |
| PhysRegion - Basin&Range     | 190 |    25 |    33 |    0.132 |    0.174 |
| PhysRegion - CO Plateau      |  54 |     3 |     7 |    0.056 |    0.130 |
| Climate - Warm-Wet           |  31 |     2 |     6 |    0.065 |    0.194 |
| Climate - Warm-Dry           |  81 |    15 |    16 |    0.185 |    0.198 |
| Climate - Cool-Wet           |  83 |     6 |    11 |    0.072 |    0.133 |
| Climate - Cool-Dry           |  49 |     5 |     7 |    0.102 |    0.143 |
| Slope - High                 | 111 |    12 |    15 |    0.108 |    0.135 |
| Slope - Low                  | 133 |    16 |    25 |    0.120 |    0.188 |

@@

|                              | n    | R^2^  | MSE    | RMSE  | MAE   | NSE   | pbias |
|------------------------------|------|-------|--------|-------|-------|-------|-------|
| Climate - Monsoon Dominated  | 3345 | 0.707 | 0.0166 | 0.129 | 0.077 | 0.694 | -12.6 |
| Climate - Snowmelt Dominated | 6152 | 0.723 | 0.017  | 0.131 | 0.094 | 0.715 | -3.7  |
| PhysRegion - Basin&Range     | 7754 | 0.749 | 0.018  | 0.133 | 0.089 | 0.741 | -5.9  |
| PhysRegion - CO Plateau      | 1743 | 0.806 | 0.013  | 0.115 | 0.081 | 0.801 | -4.4  |
| Climate - Warm-Wet           | 1613 | 0.690 | 0.015  | 0.124 | 0.081 | 0.679 | -9.9  |
| Climate - Warm-Dry           | 3134 | 0.758 | 0.023  | 0.151 | 0.097 | 0.745 | -8.1  |
| Climate - Cool-Wet           | 3134 | 0.710 | 0.013  | 0.115 | 0.085 | 0.708 | -2.5  |
| Climate - Cool-Dry           | 1616 | 0.809 | 0.014  | 0.118 | 0.082 | 0.803 | -5    |
| Slope - High                 | 4398 | 0.753 | 0.014  | 0.120 | 0.084 | 0.746 | -4.4  |
| Slope - Low                  | 5099 | 0.759 | 0.019  | 0.138 | 0.090 | 0.750 | -6.9  |

>>>>

![Map of Arizona indicating US Geological Survey (USGS) streamgages used in this study. 8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea_20241127.png){#fig-study-area}

@@

Base flow was calculated using a single-parameter, recursive digital filter technique from [@nathan1990]. This base-flow separation technique is based on a recursive digital filter used in signal analysis that separates high-frequency signals (quickflow) from low-frequency signals (base flow) [@lyne1979]. @eckhardt2023 noted that recursive digital filters lack a physical basis, but as the method is easy to automate, objective, and repeatable, it is appropriate for a regional-scale study. This technique has been used in [@ARNOLD200021; @santhi2008; @bloomfield2009; @singh2018]. It takes the form of

@@

where $b$ is base flow, $\alpha$ is the filter parameter, $Q$ is the total streamflow, and $k$ is the time step. A filter parameter $\alpha$ of 0.925 was used as in @nathan1990 and @fuka2014ecohydrology. The filter was run three times (forward, backward, forward) to attenuate the base-flow signal.

@@

![Mean annual BFI for the period of record from instrumented stream flow data.](images/BFI_Instrumented_20241126.png){#fig-instrumented-bfi}

@@

| Group                              | n   | n_pos | n_neg | perc_pos | perc_neg |
|------------------------------------|-----|-------|-------|----------|----------|
| Precipitation - Monsoon Dominated  | 87  | 8     | 21    | 0.092    | 0.241    |
| Precipitation - Snowmelt Dominated | 118 | 9     | 12    | 0.076    | 0.102    |
| PhysRegion - Basin&Range           | 156 | 14    | 26    | 0.090    | 0.167    |
| PhysRegion - CO Plateau            | 49  | 3     | 7     | 0.061    | 0.143    |
| Climate - Warm-Wet                 | 31  | 2     | 6     | 0.065    | 0.194    |
| Climate - Warm-Dry                 | 55  | 6     | 11    | 0.109    | 0.200    |
| Climate - Cool-Wet                 | 74  | 4     | 9     | 0.054    | 0.122    |
| Climate - Cool-Dry                 | 45  | 5     | 7     | 0.111    | 0.156    |
| Slope - High                       | 102 | 10    | 12    | 0.098    | 0.118    |
| Slope - Low                        | 103 | 7     | 21    | 0.068    | 0.204    |

@@

| Classification Group               | n    | R^2^  | MSE    | RMSE  | MAE   | NSE   | pbias |
|----------------|--------|--------|--------|--------|--------|--------|--------|
| Precipitation - Monsoon Dominated  | 3345 | 0.707 | 0.0166 | 0.129 | 0.077 | 0.694 | -12.6 |
| Precipitation - Snowmelt Dominated | 6152 | 0.723 | 0.017  | 0.131 | 0.094 | 0.715 | -3.7  |
| PhysRegion - Basin&Range           | 7754 | 0.749 | 0.018  | 0.133 | 0.089 | 0.741 | -5.9  |
| PhysRegion - CO Plateau            | 1743 | 0.806 | 0.013  | 0.115 | 0.081 | 0.801 | -4.4  |
| Climate - Warm-Wet                 | 1613 | 0.690 | 0.015  | 0.124 | 0.081 | 0.679 | -9.9  |
| Climate - Warm-Dry                 | 3134 | 0.758 | 0.023  | 0.151 | 0.097 | 0.745 | -8.1  |
| Climate - Cool-Wet                 | 3134 | 0.710 | 0.013  | 0.115 | 0.085 | 0.708 | -2.5  |
| Climate - Cool-Dry                 | 1616 | 0.809 | 0.014  | 0.118 | 0.082 | 0.803 | -5    |
| Slope - High                       | 4398 | 0.753 | 0.014  | 0.120 | 0.084 | 0.746 | -4.4  |
| Slope - Low                        | 5099 | 0.759 | 0.019  | 0.138 | 0.090 | 0.750 | -6.9  |
